---
title: "DATA 606 Data Project Proposal"
author: "Mia Chen"
date: "April 3, 2019"
output: html_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
drivers <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/bad-drivers/bad-drivers.csv")

head(drivers)

names(drivers) <- c("State", "Collisions", "Perc.Speeding", "Perc.Alcohol", "Perc.Not.Distracted", "Perc.No.Pre.Accident", "Insurance.Premium", "Insurance.Loss")
```



### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Are insurance incurred losses predictive of higher car insurance premiums?


### Cases 

**What are the cases, and how many are there?**

Each case represents a state in the United States. There are 51 observations in the given data set(inlcuding District of Columbia).


### Data collection 

**Describe the method of data collection.**

The data is collected from [National Highway Traffic Safety Administration](https://www-fars.nhtsa.dot.gov/Main/index.aspx) and [National Association of Insurance Commissioners](https://www.naic.org/) by [FiveThirtyEight](https://fivethirtyeight.com/features/which-state-has-the-worst-drivers/).


### Type of study 

**What type of study is this (observational/experiment)?**

This is an observational study.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

Data is collected by FiveThirtyEight and is available online here: [https://github.com/fivethirtyeight/data/tree/master/bad-drivers](https://github.com/fivethirtyeight/data/tree/master/bad-drivers). For this project, data was read from github raw csv file.



### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is car Insurance Premium and is quantitatie.


### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**

The two independent variables are State (qualitative) and Insurance Loss (quantitative).


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
library(DT)
library(psych)
library(ggplot2)

datatable(drivers)
```

```{r}
describe(drivers$Insurance.Premium)
describe(drivers$Insurance.Loss)
summary(drivers)
```

```{r}
ggplot(drivers, aes(x=Insurance.Loss)) + geom_histogram(binwidth = 10)
```

```{r}
ggplot(drivers, aes(x=Insurance.Premium)) + geom_histogram(binwidth = 80)
```

